---
title: "Regression in uwIntroStats"
date: "22 July, 2015"
output: html_document
---

In the `uwIntroStats` package, we have set out to make regression and analysis easier by:

* allowing the user to specify any type of regression from one function
* allowing the user to specify multiple-partial F-tests
* displaying output in a more intuitive fashion than base R

This function is `regress()`. The basic arguments to this function, which unlock all of its potential, are

* `fnctl` - the functional
* `formula` - the formula for the linear model
* `data` - the data to use for the model
* `id` - the identification variable in data with repeated measurements

We use the concept of a *functional* to handle our first goal. A functional takes a function as its argument and returns a number - hence the mean is a functional, because it takes a distribution as its argument and returns a single number. The allowed functionals to `regress()` are

|Functional | Type of Regression | Previous command (package)|
|---|:---|:---|
|`"mean"` | Linear Regression | `lm()` (`stats` - base R) |
|`"geometric mean"` | Linear Regression on logarithmically transformed Y | `lm()`, with Y log transformed (`stats` - base R) |
|`"odds"` | Logistic Regression | `glm(family = binomial)` (`stats` - base R) |
|`"rate"` | Poisson Regression | `glm(family = poisson)` (`stats` - base R) |
|`"hazard"` | Proportional Hazards Regression | `coxph()` (`survival`)|

The *formula* to `regress()` is the same as a formula given to `lm()` or any of the other regression commands from base R, `survival`, or `geepack`, but with one small addition. To address our second goal of allowing the user to specify multiple-partial F-tests, we have added a special function - `U()` - which can be added to the formula. The `U()` function is documented more fully in ["User Specified Multiple-Partial F-tests in Regression"](u_intro.html).

The *data* argument is exactly the same as that in `lm()` or any of the other regression commands.

Last, *id* allows the user to fit a generalized estimating equations (GEE) model while using the same syntax as any of the functionals to `regress()`. The GEE framework is a useful way to model correlated data, which often comes in the form of repeated measurements.

# Linear Regression
As a first example, we run a linear regression analysis of atrophy on age and male, from the `mri` data.
```{r}
## Preparing our R session
library(uwIntroStats)
data(mri)
```
```{r}
regress("mean", atrophy ~ age + male, data = mri)
```

This call automatically prints the coefficients table. First, notice that by default robust standard error estimates (calculated using the `sandwich` package) are returned, in addition to the naive estimates. The robust estimates are also used to perform the inference - thus the confidence intervals, statistics, and p-values use these estimates of the standard error.

If we did not use robust standard error estimates, then in the case of linear regression we would be assuming that all groups have the same variance. Then any inference we make could be on the fact that the variances are different, rather than just on the means - which is usually what we want in linear regression. 

F-statistics are also displayed by default. 